---
title: "hms_520_TB_Final"
author: "Sophie Whikehart and Ye Htet Naing"
date: "2024-12-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install libraries
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Final Project - Global Mortality and Risk Factor Contributions for Tuberculosis Estimates from 2015 and 2020

## Abstract

- Ye

- Give an overview of the project, what our goals are and what we hope to find 

## Introduction

- Ye

Provide a brief background: what do we already know about this topic? Why is it interesting? What do you want to add to what already exists?

---

## Data Explanation

## Data Description

For this project we are using the data set from IHME titled: Global Burden of Disease 2021 [GBD 2021] Tuberculosis Estimates 1990-2021 

This data set includes estimates of burden associated with all-form tuberculosis for GBD countries between 1990 and 2021. Tuberculosis mortality was informed by vital registration, verbal autopsy, sample-based vital registration and mortality surveillance data. TB morbidity data includes annual case notifications, data from prevalence surveys, and estimated cause specific mortality [CSMR] of TB among HIV-positive and HIV-negative individuals (IHME GBD 2021).

For our project we are utilizing the `IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX` which contains risk deleted deaths due to all-form tuberculosis for alcohol use, smoking, and diabetes and all three risk factors combined by adult age groups by country for 2015, 2020 and 2021. 

## Methods 

### Data pre-processing 

```{r}

# read in the data 
mortality_data <- read_excel("/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/data/IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX")

## check for missing values 
colSums(is.na(mortality_data))
```

1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

```{r}

# filter for data just on region and super regions 

region_data <- mortality_data %>%
  filter(location_type %in% c("region"))

# data wrangling -- calculate percent change in mortality by regions 

mortality_region <- region_data %>%
  mutate(mortality_change = ((mort_2020_count_mean - mort_2015_count_mean) / mort_2015_count_mean) * 100) %>%
  group_by(age_group_name, location_name) %>%
  summarize(mean_mortality_change = mean(mortality_change, na.rm = TRUE))

# visualization on region based on decreasing in mortality 

## highlighting the largest reduction

ggplot(mortality_region, aes(x = reorder(location_name, -mean_mortality_change), y = mean_mortality_change)) + 
  geom_point(color = "red", size = 3) + 
  geom_segment(aes(xend = location_name, yend = 0), color = "black") +  # Lines from points to zero
  coord_flip() +  # Flip axes for horizontal bars
  labs(title = "Mean Mortality Change from 2015 - 2020 (All Ages)",
       x = "Region",
       y = "Mean Mortality Change") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

## Regions with the largest negative values (located at the top of the plot) have seen the biggest reduction in mortality during the 2015-2020 period. These are likely areas where health interventions, improvements in healthcare, or other factors might have led to a reduction in deaths.

## Regions with positive values will show increased mortality, suggesting areas where the mortality rate has risen, potentially due to challenges like poor healthcare access, economic difficulties, or other systemic issues.

## plot provides a comparative view of how different regions have performed in terms of mortality over the 2015-2020 period. This could help identify patterns, trends, or regions that need additional focus in terms of health interventions or resources.

## If you are analyzing regional disparities, the plot helps identify which regions have been improving (decreasing mortality) versus those that might require attention due to rising mortality rates.

# Create a line plot to visualize trends by age group
ggplot(mortality_region, aes(x = reorder(location_name, mean_mortality_change), 
                             y = mean_mortality_change, 
                             color = age_group_name, 
                             group = age_group_name)) +
  geom_line() +  # Adds lines for each age group
  geom_point() +  # Adds points at each location for clarity
  coord_flip() +  # Flip coordinates for horizontal lines
  labs(title = "Trends in Mean Mortality Change by Location",
       x = "Location",
       y = "Mean Mortality Change") +
  theme_minimal() +  # Clean minimal theme
  theme(axis.text.y = element_text(size = 10)) +  # Adjust text size for readability
  scale_color_brewer(palette = "Set2")  # Use a color palette for age groups

## represents average change in mortality rates for each location
  ## positive value indicates increase in mortality while negative value indicates decrease

## lines and points corresponding to each age group show how mean mortality change trend varies for that specific group across different locations

## slope of each line indicates direction and strength of change 
  ## if line is sloped upward it indicates that mortality has increased in those locations for that age group over time 
```

2. What is the relative contribution of different risk factors [ex - smoking, alcohol, and diabetes] to mortality?



3. Do regions or age groups with higher mortality reductions also show lower risk factor contributions?

 

---

## Results 

- Ye & Sophie 



