---
title: "HMS_520_TB_Final"
author: "Sophie Whikehart and Ye Htet Naing"
date: "2024-12-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

# install libraries
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Final Project - Global Mortality and Risk Factor Contributions for Tuberculosis Estimates from 2015 and 2020

## Abstract

The World Health Organization (WHO) End TB Strategy set ambitious targets to reduce tuberculosis (TB) mortality by 35% and incidence by 20% between 2015 and 2020. While global progress has been reported, age-specific and regional analyses remain limited. This study utilizes the Global Burden of Disease 2021 (GBD 2021) Tuberculosis Estimates to assess trends in TB mortality and the contributions of smoking, alcohol use, and diabetes as risk factors.

Our analysis examines TB mortality trends from 2015 to 2020 across age groups, regions, and risk factors using mortality counts from the GBD data set. 

## Introduction

Tuberculosis (TB) is an ancient infectious disease with evidence of bone in TB in Egyptian mummies (2400 BC). TB is also known as consumption, the white plague, phthisis and the graveyard cough. TB is an airborne infection that is spread from someone with active TB. 1/10 people on average who are infected will go on to develop active TB. On the other hand, if you have latent TB you have no symptoms and cannot spread TB infection to others.

Testing for latent TB includes tuberculin skin tests and interferon gamma release assays [blood tests]. The sites of active TB disease include the lungs [pulmonary TB] and can also involve other organs such as the bones and intestines. TB diagnosis is done with specimen collection of sputum, CSF or pleural fluid. The gold standard for diagnosis is through a culture but the disadvantage it that there is long turnaround time [up to 6-8 weeks]. Other methods include a smear microscopy and molecular methods such as Gene Xpert.

TB data sources often come from prevalence surveys, surveys measuring latent TB infection, TB case notifications, vital registration data and verbal autopsy data. However, data sources are imperfect and data quality and comparability can affect the measurement of the global burden of TB. 

Tuberculosis (TB) continues to be one of the leading causes of infectious disease mortality globally, posing significant challenges to public health systems. In response, the World Health Organization (WHO) launched the End TB Strategy, aiming to reduce TB mortality by 35% and TB incidence by 20% between 2015 and 2020. While global progress has been documented, there is a limited understanding of how these trends vary by age and region, as well as the role of modifiable risk factors such as smoking, alcohol use, and diabetes. Age-specific evaluations are critical to inform tailored interventions and ensure equitable progress across populations.

In this study, we leverage the Global Burden of Disease 2021 (GBD 2021) Tuberculosis Estimates to examine TB mortality trends from 2015 to 2020 across different age groups and regions. The data set includes mortality counts stratified by location, year, and age, as well as estimates of deaths attributable to smoking, alcohol use, and diabetes.

Our primary research questions include:

1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

2. What is the relative contribution of different risk factors (e.g., smoking, alcohol use, and diabetes) to TB mortality in 2015 and 2020?

3. Do regions or age groups with higher reductions in TB mortality also show lower contributions of risk factors?

By addressing these questions, this project seeks to fill critical gaps in understanding TB mortality trends and their driving factors. Our findings will provide evidence to guide targeted interventions and strengthen global efforts toward achieving the WHO End TB Strategy goals.


## Data Description

For this project we are using the data set from IHME titled: Global Burden of Disease 2021 [GBD 2021] Tuberculosis Estimates 1990-2021 

This data set includes estimates of burden associated with all-form tuberculosis for GBD countries between 1990 and 2021. Tuberculosis mortality was informed by vital registration, verbal autopsy, sample-based vital registration and mortality surveillance data. TB morbidity data includes annual case notifications, data from prevalence surveys, and estimated cause specific mortality [CSMR] of TB among HIV-positive and HIV-negative individuals (IHME GBD 2021).

For our project we are utilizing the `IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX` which contains risk deleted deaths due to all-form tuberculosis for alcohol use, smoking, and diabetes and all three risk factors combined by adult age groups by country for 2015, 2020 and 2021. 

## Methods 

The variables of interest in our project were the mortality counts, risk factors and geographic and age group stratification. We pre-processed the data by:

1. Importing data set and filtering relevant columns
2. Checking and address missing data
3. Creating calculated variables 
  - Percent change in mortality between 2015 and 2020 
  - Proportional contribution of each risk factor to attributable TB deaths 
  
We also conducted the following analyses 

1. Trend Analysis 
  - Calculate percent change in mortality by region and age group from 2015 - 2020 
  - Visualize the trend using bar plot and line plot to highlight the change in mortality 
  
2. Attributable Risk Factor Analysis 
  - Summarize the contribution of smoking, alcohol use and diabetes to TB mortality across age groups and regions 
  - Use grouped bar plots to visualize attributable mortality by risk factor 
  
3. Association Between Mortality Reduction and Risk Factor Contribution 
  - Assess relationship between reduction in mortality rate and average risk factor contribution using linear regression model
  - Visualize the association with scatter plot and regression line 
  
```{r, include = FALSE}

# read in the data 
mortality_data <- read_excel("/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/data/IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX")

## filter for relevant columns 
mortality_data <- mortality_data %>%
  select(
    location_name,
    location_type,
    age_group_name,
    mort_2015_count_mean,
    mort_2020_count_mean,
    rmv_mean_smoking,
    rmv_mean_alcohol,
    rmv_mean_diabetes,
    rmv_mean_all_risk
  )

## check for missing values 
colSums(is.na(mortality_data))
```

# Question 1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

```{r}

# filter for data just on region and super regions 

region_data <- mortality_data %>%
  filter(location_type %in% c("region"))

# data wrangling -- calculate percent change in mortality by regions 

mortality_region <- region_data %>%
  mutate(mortality_change = ((mort_2020_count_mean - mort_2015_count_mean) / mort_2015_count_mean) * 100) %>%
  group_by(age_group_name, location_name) %>%
  summarize(mean_mortality_change = mean(mortality_change, na.rm = TRUE))

# visualization on region based on decreasing in mortality 

## highlighting the largest reduction

fig_1 <- ggplot(mortality_region, aes(x = reorder(location_name, -mean_mortality_change), y = mean_mortality_change)) + 
  geom_point(color = "red", size = 3) + 
  geom_segment(aes(xend = location_name, yend = 0), color = "black") +  # Lines from points to zero
  coord_flip() +  # Flip axes for horizontal bars
  labs(title = "Fig 1. Mean Mortality Change from 2015 - 2020 (All Ages)",
       x = "Region",
       y = "Mean Mortality Change") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig_1

## Regions with the largest negative values (located at the top of the plot) have seen the biggest reduction in mortality during the 2015-2020 period. These are likely areas where health interventions, improvements in healthcare, or other factors might have led to a reduction in deaths.

## Regions with positive values will show increased mortality, suggesting areas where the mortality rate has risen, potentially due to challenges like poor healthcare access, economic difficulties, or other systemic issues.

## plot provides a comparative view of how different regions have performed in terms of mortality over the 2015-2020 period. This could help identify patterns, trends, or regions that need additional focus in terms of health interventions or resources.

## If you are analyzing regional disparities, the plot helps identify which regions have been improving (decreasing mortality) versus those that might require attention due to rising mortality rates.

# Create a line plot to visualize trends by age group
fig_2 <- ggplot(mortality_region, aes(x = reorder(location_name, -mean_mortality_change), 
                             y = mean_mortality_change, 
                             color = age_group_name, 
                             group = age_group_name)) +
  geom_line() +  # Adds lines for each age group
  geom_point() +  # Adds points at each location for clarity
  coord_flip() +  # Flip coordinates for horizontal lines
  labs(title = "Fig 2. Trends in Mean Mortality Change by Location",
       x = "Location",
       y = "Mean Mortality Change") +
  theme_minimal() +  # Clean minimal theme
  theme(axis.text.y = element_text(size = 10)) +  # Adjust text size for readability
  scale_color_brewer(palette = "Set2")  # Use a color palette for age groups

fig_2

## represents average change in mortality rates for each location
  ## positive value indicates increase in mortality while negative value indicates decrease

## lines and points corresponding to each age group show how mean mortality change trend varies for that specific group across different locations

## slope of each line indicates direction and strength of change 
  ## if line is sloped upward it indicates that mortality has increased in those locations for that age group over time 
```

# Question 2. What is the relative contribution of different risk factors [ex - smoking, alcohol, and diabetes] to mortality?

```{r}

# Calculate attributable mortality 

risk_factors <- mortality_data %>%
  mutate(
    prop_smoking = rmv_mean_smoking / rmv_mean_all_risk,
    prop_alcohol = rmv_mean_alcohol / rmv_mean_all_risk,
    prop_diabetes = rmv_mean_diabetes / rmv_mean_all_risk
  )

## this computes proportion of total attributable mortality due to each risk factor

# summarize by age 

summary_by_age <- risk_factors %>%
  group_by(age_group_name) %>%
  summarise(
    smoking_mortality = sum(rmv_mean_smoking, na.rm = TRUE),
    alcohol_mortality = sum(rmv_mean_alcohol, na.rm = TRUE),
    diabetes_mortality = sum(rmv_mean_diabetes, na.rm = TRUE),
    total_attributable = sum(rmv_mean_all_risk, na.rm = TRUE)
  )

# summarize_by_region <- risk_factors %>%


# visualize 

summary_long <- summary_by_age %>%
  pivot_longer(
    cols = c(smoking_mortality, alcohol_mortality, diabetes_mortality),
    names_to = "risk_factor",
    values_to = "mortality"
  )

fig_3 <- ggplot(summary_long, aes(x = age_group_name, y = mortality, fill = risk_factor)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Fig 3. Attributable Mortality by Age Group",
    x = "Age Group",
    y = "Attributable Deaths",
    fill = "Risk Factor"
  ) +
  theme_minimal()

fig_3
```

```{r}
summarize_by_region <- risk_factors %>%
  filter(location_type %in% c("superregion", "region")) %>%  # Filter for superregion and region
  group_by(location_name, location_type) %>%  # Group by region and type
  summarise(
    smoking_mortality = sum(rmv_mean_smoking, na.rm = TRUE),
    alcohol_mortality = sum(rmv_mean_alcohol, na.rm = TRUE),
    diabetes_mortality = sum(rmv_mean_diabetes, na.rm = TRUE),
    total_attributable = sum(rmv_mean_all_risk, na.rm = TRUE)
  )

# Reshape to long format
region_long <- summarize_by_region %>%
  pivot_longer(
    cols = c(smoking_mortality, alcohol_mortality, diabetes_mortality),
    names_to = "risk_factor",
    values_to = "mortality"
  )

fig_4 <- ggplot(region_long, aes(x = location_name, y = mortality, fill = risk_factor)) +
  geom_bar(stat = "identity", position = "dodge") +  # Grouped bars
  labs(
    title = "Fig 4. Attributable Mortality by Region",
    x = "Region",
    y = "Mortality",
    fill = "Risk Factor"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom"
  )

fig_4
```

# Question 3. Do regions or age groups with higher mortality reductions also show lower risk factor contributions?

```{r}

# calculate mortality reduction 

mortality_data <- mortality_data %>%
  mutate (
    reduction_rate = ((mort_2015_count_mean - mort_2020_count_mean) / mort_2015_count_mean) * 100
  )

# summary risk factor contribution 

mortality_data <- mortality_data %>%
  rowwise() %>%
  mutate(
    avg_risk_factor_contribution = mean(c(rmv_mean_smoking, rmv_mean_alcohol, rmv_mean_diabetes, rmv_mean_all_risk), na.rm = TRUE)
  )

# linear regression 

model <- lm(avg_risk_factor_contribution ~ reduction_rate, data = mortality_data)

# view model 
summary(model)

# visualize 
fig_5 <- ggplot(mortality_data, aes(x = reduction_rate, y = avg_risk_factor_contribution)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Fig 5. Linear Regression: Mortality Reduction vs. Risk Factor Contributions",
    x = "Mortality Reduction Rate (%)",
    y = "Average Risk Factor Contribution"
  ) +
  theme_minimal()

fig_5
```

## Results 

In our first question we calculated and visualized percent changes in TB mortality across regions and age groups. Our results show that there are differences in reductions and increases in TB mortality by region and age group. In **Figure 1.** we calculated the mean mortality change from 2015 - 2020 by all ages and grouped by region and super region. The results of this figure show the visualization of region based on decreasing in mortality. Therefore Eastern Europe, Central Europe and Central Asia have seen the biggest reduction in mortality from 2015 to 2020. There are likely the areas where health interventions, improvements in healthcare or other factors have led to a reduction in deaths due to TB. On the other hand regions such as High-income North America, Oceania and Central Latin America have shown the highest increase in mortality. This suggests that these areas are where mortality rates have risk which could be due to potential challenges such as poor healthcare access, economic difficulty and other systemic issues. This graph helps identify what regions need additional focus in terms of health interventions or resources to help reduce TB mortality. 

**Figure 2.** shows the trends in mean mortality change by region and super region but this time it is grouped by age groups. This figure highlights that consistently we see an increase of mortality change by the age groups in the 70+ and 50-69 year age group. Because these age groups consistently exhibit higher mortality change across multiple locations it indicates that there is need for age-specific interventions for older populations in all regions. 

Our second question wanted to provide insight into the relative contributions of different risk factors (smoking, alcohol and diabetes) to mortality both by age group as well as by region. In **Figure 3.** the bar plot shows how attributable mortality for each risk factor varies across different age groups. As we can see from the graph in ages 15 - 49 the risk factor for diabetes is the highest with smoking slightly higher than alcohol mortality. In the age group of 50 - 69 alcohol is now the highest with diabetes second and smoking third. In 70+ age group, alcohol remains the highest with smoking second and diabetes third. Finally, in all ages alcohol is the highest with diabetes in second and smoking mortality third. 

Therefore in ages 15-49 with diabetes being the highest contribute there could be a need for early screening and prevention programs for diabetes and promoting lifestyle modifications among younger populations. For ages 50-69 with alcohol being the leading risk factor in this age group interventions could include strengthening alcohol education and harm reduction initiatives as well as enhancing access to treatment for alcohol dependency. For ages 70+ alcohol remains the highest contributor which suggests long-term impact of alcohol use or age related vulnerabilities but smoking is the second highest so interventions could include the need for lifelong smoking cessation programs and addressing comorbidities related to alcohol and smoking in older adults. 

**Figure 4** shows the attributable mortality risk by region to see which regions have the highest morbidity due to different risk factors. From this figure we can see that regions with highest risk factor attributable deaths are Sub-Saharan Africa, South Asia and Eastern Sub-Saharan Africa. These likely reflect broader challenges such as limited access to healthcare and preventative services, socioeconomic inequalities that exacerbate exposure to risk factors and insufficient infrastructure for managing chronic diseases and addiction. In order to reduce these risk factors, public health policy and intervention priorities should be placed on these regions.

Finally our last question wanted to see if regions or ages groups with higher mortality reductions also showed lower risk factor contributions? We calculated the mortality reduction rate and then ran a linear regression model to provide insights about the relationship between mortality reduction and average risk factor contribution. The slope B1 was at -25.88 which indicates a very slight and non-significant decrease in average risk factor contribution for each 1% increase in mortality reduction rate. The p-value for the slop was 0.82 which indicates that there is no statistically significant relationship between mortality reduction and risk factor contribution. In addition the R^2 value was at 5.629 X 10^-5 which indicates that the reduction rate explains virtually none of the variation in the average risk factor contribution. The residuals also range quite widely to show that the data points are spread out and the model does not capture much variation. 

In **Figure 5.** we visualized the results of our linear regression model and you can clearly see that there is no correlation. Because there is no significant relationship it is most likely that mortality reductions may be driven by other factors such as improvements in healthcare systems or better access to treatments rather than direct reductions in risk factor contributions. This highlights the need for broader exploration of mortality determinants beyond the examined risk factors. 

# Conclusion 

This project aimed to assess the trends in tuberculosis mortality from 2015 - 2020 and examine the contribution of smoking, alcohol use and diabetes as risk factors. Our analysis revealed several important findings 

1. TB Mortality Trends 
2. Risk Factor Contribution
3. Association between Mortality Reduction and Risk Factor Contribution

These findings show the importance of addressing both the disease itself and modifiable risk factors contributing to TB mortality. Targeted interventions based on the findings from this report could potentially further accelerate progress toward the WHO End TB Strategy's mortality reduction goals. Additionally region and age specific approaches tailored to local health challenges are crucial to achieve sustained reductions in TB mortality globally. 

```{r, include = FALSE}

# save figures 

# Save the plot
ggsave(
  filename = "/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/figures/fig_1.jpg",
  plot = fig_1
)

ggsave(
  filename = "/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/figures/fig_2.jpg",
  plot = fig_2
)

ggsave(
  filename = "/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/figures/fig_3.jpg",
  plot = fig_3
)

ggsave(
  filename = "/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/figures/fig_4.jpg",
  plot = fig_4
)

ggsave(
  filename = "/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/figures/fig_5.jpg",
  plot = fig_5
)
```





