---
title: "hms_520_TB_Final"
author: "Sophie Whikehart and Ye Htet Naing"
date: "2024-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install libraries
library(readxl)
library(dplyr)
library(tidyr)
```

# Final Project - Global Mortality and Risk Factor Contributions for Tuberculosis Estimates from 2015 and 2020

## Abstract

- Ye

- Give an overview of the project, what our goals are and what we hope to find 

## Introduction

- Ye

Provide a brief background: what do we already know about this topic? Why is it interesting? What do you want to add to what already exists?

---

## Data Explanation

## Data Description

For this project we are using the data set from IHME titled: Global Burden of Disease 2021 [GBD 2021] Tuberculosis Estimates 1990-2021 

This data set includes estimates of burden associated with all-form tuberculosis for GBD countries between 1990 and 2021. Tuberculosis mortality was informed by vital registration, verbal autopsy, sample-based vital registration and mortality surveillance data. TB morbidity data includes annual case notifications, data from prevalence surveys, and estimated cause specific mortality [CSMR] of TB among HIV-positive and HIV-negative individuals (IHME GBD 2021).

For our project we are utilizing the `IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX` which contains risk deleted deaths due to all-form tuberculosis for alcohol use, smoking, and diabetes and all three risk factors combined by adult age groups by country for 2015, 2020 and 2021. 

## Methods 

### Data pre-processing 

```{r}

# read in the data 
mortality_data <- read_excel("/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/data/IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX")

## check for missing values 
colSums(is.na(mortality_data))

## rename columns 

cleaned_data <- mortality_data %>%
  rename(
    location = location_name,
    location_type = location_type,
    age_group = age_group_name,
    mort_2015 = mort_2015_count_mean,
    mort_2020 = mort_2020_count_mean,
    smoking_risk = rmv_mean_smoking,
    alcohol_risk = rmv_mean_alcohol,
    diabetes_risk = rmv_mean_diabetes,
    total_risk = rmv_mean_all_risk
  )
```

1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

2. What is the relative contribution of different risk factors [ex - smoking, alcohol, and diabetes] to mortality in 2015 and 2020?

3. Do regions or age groups with higher mortality reductions also show lower risk factor contributions?

 

---

## Results 

- Ye & Sophie 



