---
title: "HMS_520_TB_Final"
author: "Sophie Whikehart and Ye Htet Naing"
date: "2024-12-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

# install libraries
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Final Project - Global Mortality and Risk Factor Contributions for Tuberculosis Estimates from 2015 and 2020

## Abstract

The World Health Organization (WHO) End TB Strategy set ambitious targets to reduce tuberculosis (TB) mortality by 35% and incidence by 20% between 2015 and 2020. While global progress has been reported, age-specific and regional analyses remain limited. This study utilizes the Global Burden of Disease 2021 (GBD 2021) Tuberculosis Estimates to assess trends in TB mortality and the contributions of smoking, alcohol use, and diabetes as risk factors.

Our analysis examines TB mortality trends from 2015 to 2020 across age groups, regions, and risk factors using mortality counts from the GBD data set. 

## Introduction

Tuberculosis (TB) continues to be one of the leading causes of infectious disease mortality globally, posing significant challenges to public health systems. In response, the World Health Organization (WHO) launched the End TB Strategy, aiming to reduce TB mortality by 35% and TB incidence by 20% between 2015 and 2020. While global progress has been documented, there is a limited understanding of how these trends vary by age and region, as well as the role of modifiable risk factors such as smoking, alcohol use, and diabetes. Age-specific evaluations are critical to inform tailored interventions and ensure equitable progress across populations.

In this study, we leverage the Global Burden of Disease 2021 (GBD 2021) Tuberculosis Estimates to examine TB mortality trends from 2015 to 2020 across different age groups and regions. The data set includes mortality counts stratified by location, year, and age, as well as estimates of deaths attributable to smoking, alcohol use, and diabetes.

Our primary research questions include:

1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

2. What is the relative contribution of different risk factors (e.g., smoking, alcohol use, and diabetes) to TB mortality in 2015 and 2020?

3. Do regions or age groups with higher reductions in TB mortality also show lower contributions of risk factors?

By addressing these questions, this project seeks to fill critical gaps in understanding TB mortality trends and their driving factors. Our findings will provide evidence to guide targeted interventions and strengthen global efforts toward achieving the WHO End TB Strategy goals.


## Data Description

For this project we are using the data set from IHME titled: Global Burden of Disease 2021 [GBD 2021] Tuberculosis Estimates 1990-2021 

This data set includes estimates of burden associated with all-form tuberculosis for GBD countries between 1990 and 2021. Tuberculosis mortality was informed by vital registration, verbal autopsy, sample-based vital registration and mortality surveillance data. TB morbidity data includes annual case notifications, data from prevalence surveys, and estimated cause specific mortality [CSMR] of TB among HIV-positive and HIV-negative individuals (IHME GBD 2021).

For our project we are utilizing the `IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX` which contains risk deleted deaths due to all-form tuberculosis for alcohol use, smoking, and diabetes and all three risk factors combined by adult age groups by country for 2015, 2020 and 2021. 

## Methods 

The variables of interest in our project were the mortality counts, risk factors and geographic and age group stratification. We pre-processed the data by:

1. Importing data set and filtering relevant columns
2. Checking and address missing data
3. Creating calculated variables 
  - Percent change in mortality between 2015 and 2020 
  - Proportional contribution of each risk factor to attributable TB deaths 
  
We also conducted the following analyses 

1. Trend Analysis 
  - Calculate percent change in mortality by region and age group from 2015 - 2020 
  - Visualize the trend using bar plot and line plot to highlight the change in mortality 
  
2. Attributable Risk Factor Analysis 
  - Summarize the contribution of smoking, alcohol use and diabetes to TB mortality across age groups and regions 
  - Use grouped bar plots to visualize attributable mortality by risk factor 
  
3. Association Between Mortality Reduction and Risk Factor Contribution 
  - Assess relationship between reduction in mortality rate and average risk factor contribution using linear regression model
  - Visualize the association with scatter plot and regression line 
  
```{r, include = FALSE}

# read in the data 
mortality_data <- read_excel("/Users/seanwilcox/Desktop/HMS_520_Final_Project_TB/data/IHME_GBD_2021_TB_MORTALITY_RISK_Y2024M03D19.XLSX")

## filter for relevant columns 
mortality_data <- mortality_data %>%
  select(
    location_name,
    location_type,
    age_group_name,
    mort_2015_count_mean,
    mort_2020_count_mean,
    rmv_mean_smoking,
    rmv_mean_alcohol,
    rmv_mean_diabetes,
    rmv_mean_all_risk
  )

## check for missing values 
colSums(is.na(mortality_data))
```

---

# Question 1. How have mortality rates changed from 2015 to 2020 across different age groups and regions?

```{r}

# filter for data just on region and super regions 

region_data <- mortality_data %>%
  filter(location_type %in% c("region"))

# data wrangling -- calculate percent change in mortality by regions 

mortality_region <- region_data %>%
  mutate(mortality_change = ((mort_2020_count_mean - mort_2015_count_mean) / mort_2015_count_mean) * 100) %>%
  group_by(age_group_name, location_name) %>%
  summarize(mean_mortality_change = mean(mortality_change, na.rm = TRUE))

# visualization on region based on decreasing in mortality 

## highlighting the largest reduction

ggplot(mortality_region, aes(x = reorder(location_name, -mean_mortality_change), y = mean_mortality_change)) + 
  geom_point(color = "red", size = 3) + 
  geom_segment(aes(xend = location_name, yend = 0), color = "black") +  # Lines from points to zero
  coord_flip() +  # Flip axes for horizontal bars
  labs(title = "Fig 1. Mean Mortality Change from 2015 - 2020 (All Ages)",
       x = "Region",
       y = "Mean Mortality Change") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

## Regions with the largest negative values (located at the top of the plot) have seen the biggest reduction in mortality during the 2015-2020 period. These are likely areas where health interventions, improvements in healthcare, or other factors might have led to a reduction in deaths.

## Regions with positive values will show increased mortality, suggesting areas where the mortality rate has risen, potentially due to challenges like poor healthcare access, economic difficulties, or other systemic issues.

## plot provides a comparative view of how different regions have performed in terms of mortality over the 2015-2020 period. This could help identify patterns, trends, or regions that need additional focus in terms of health interventions or resources.

## If you are analyzing regional disparities, the plot helps identify which regions have been improving (decreasing mortality) versus those that might require attention due to rising mortality rates.

# Create a line plot to visualize trends by age group
ggplot(mortality_region, aes(x = reorder(location_name, mean_mortality_change), 
                             y = mean_mortality_change, 
                             color = age_group_name, 
                             group = age_group_name)) +
  geom_line() +  # Adds lines for each age group
  geom_point() +  # Adds points at each location for clarity
  coord_flip() +  # Flip coordinates for horizontal lines
  labs(title = "Fig 2. Trends in Mean Mortality Change by Location",
       x = "Location",
       y = "Mean Mortality Change") +
  theme_minimal() +  # Clean minimal theme
  theme(axis.text.y = element_text(size = 10)) +  # Adjust text size for readability
  scale_color_brewer(palette = "Set2")  # Use a color palette for age groups

## represents average change in mortality rates for each location
  ## positive value indicates increase in mortality while negative value indicates decrease

## lines and points corresponding to each age group show how mean mortality change trend varies for that specific group across different locations

## slope of each line indicates direction and strength of change 
  ## if line is sloped upward it indicates that mortality has increased in those locations for that age group over time 
```

---

# Question 2. What is the relative contribution of different risk factors [ex - smoking, alcohol, and diabetes] to mortality?

```{r}

# Calculate attributable mortality 

risk_factors <- mortality_data %>%
  mutate(
    prop_smoking = rmv_mean_smoking / rmv_mean_all_risk,
    prop_alcohol = rmv_mean_alcohol / rmv_mean_all_risk,
    prop_diabetes = rmv_mean_diabetes / rmv_mean_all_risk
  )

## this computes proportion of total attributable mortality due to each risk factor

# summarize by age 

summary_by_age <- risk_factors %>%
  group_by(age_group_name) %>%
  summarise(
    smoking_mortality = sum(rmv_mean_smoking, na.rm = TRUE),
    alcohol_mortality = sum(rmv_mean_alcohol, na.rm = TRUE),
    diabetes_mortality = sum(rmv_mean_diabetes, na.rm = TRUE),
    total_attributable = sum(rmv_mean_all_risk, na.rm = TRUE)
  )

# summarize_by_region <- risk_factors %>%


# visualize 

summary_long <- summary_by_age %>%
  pivot_longer(
    cols = c(smoking_mortality, alcohol_mortality, diabetes_mortality),
    names_to = "risk_factor",
    values_to = "mortality"
  )

ggplot(summary_long, aes(x = age_group_name, y = mortality, fill = risk_factor)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Fig 3. Attributable Mortality by Age Group",
    x = "Age Group",
    y = "Attributable Deaths",
    fill = "Risk Factor"
  ) +
  theme_minimal()
```
```{r}
summarize_by_region <- risk_factors %>%
  filter(location_type %in% c("superregion", "region")) %>%  # Filter for superregion and region
  group_by(location_name, location_type) %>%  # Group by region and type
  summarise(
    smoking_mortality = sum(rmv_mean_smoking, na.rm = TRUE),
    alcohol_mortality = sum(rmv_mean_alcohol, na.rm = TRUE),
    diabetes_mortality = sum(rmv_mean_diabetes, na.rm = TRUE),
    total_attributable = sum(rmv_mean_all_risk, na.rm = TRUE)
  )

# Reshape to long format
region_long <- summarize_by_region %>%
  pivot_longer(
    cols = c(smoking_mortality, alcohol_mortality, diabetes_mortality),
    names_to = "risk_factor",
    values_to = "mortality"
  )

ggplot(region_long, aes(x = location_name, y = mortality, fill = risk_factor)) +
  geom_bar(stat = "identity", position = "dodge") +  # Grouped bars
  labs(
    title = "Fig 4. Attributable Mortality by Region",
    x = "Region",
    y = "Mortality",
    fill = "Risk Factor"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom"
  )
```

---

# Question 3. Do regions or age groups with higher mortality reductions also show lower risk factor contributions?

```{r}

# calculate mortality reduction 

mortality_data <- mortality_data %>%
  mutate (
    reduction_rate = ((mort_2015_count_mean - mort_2020_count_mean) / mort_2015_count_mean) * 100
  )

# summary risk factor contribution 

mortality_data <- mortality_data %>%
  rowwise() %>%
  mutate(
    avg_risk_factor_contribution = mean(c(rmv_mean_smoking, rmv_mean_alcohol, rmv_mean_diabetes, rmv_mean_all_risk), na.rm = TRUE)
  )

# linear regression 

model <- lm(avg_risk_factor_contribution ~ reduction_rate, data = mortality_data)

# view model 
summary(model)

# visualize 
ggplot(mortality_data, aes(x = reduction_rate, y = avg_risk_factor_contribution)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Fig 5. Linear Regression: Mortality Reduction vs. Risk Factor Contributions",
    x = "Mortality Reduction Rate (%)",
    y = "Average Risk Factor Contribution"
  ) +
  theme_minimal()
```

---

## Results 

- Ye & Sophie 



